A Constructive Ordered Predicate Logic for Category Theory

### 2022-09-21 paper/paper.tex
intro tweaks, start working on shortening the syntax section

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\ohol}{OHol}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

\newcommand{\paramPresheaf}[1]{\mathcal P^{#1}}
\newcommand{\pmPresheaf}{\paramPresheaf\pm}
\newcommand{\posPresheaf}{\paramPresheaf+}
\newcommand{\negPresheaf}{\paramPresheaf-}
\newcommand{\negPresheafApp}[2]{#1\,#2}
\newcommand{\negPresheafAppPtoX}[2]{#2\,#1}
\newcommand{\posPresheafApp}[2]{#1\,#2}

@@

\title{An Ordered Logic for Category Theory}
\author{Max S. New\inst{1,2} and Daniel R. Licata\inst{2}}
\institute{University of Michigan \and Wesleyan University}

@@

  We develop a sound and complete notion of categorical model and give
  a model construction that shows that the theorems are valid in
  enriched and internal categories. This means that while the proofs
  in our type theory look like standard set-based arguments, the
  substructural syntactic restrictions ensure that the proofs carry
  over to enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that these categorical structures were common in
logic and computer science, and that formal systems like logics, type
theories and programming languages can be modeled by sound and
complete semantic models given by notions of structured
category\cite{lawvere,goguen,lambek,moggi}, a phenomenon sometimes
referred to as the Curry-Howard-Lambek correspondence
\cite{curry-howard-lambek}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research known as
categorical logic, where
new type theories are designed based directly on category-theoretic
semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in pure mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

@@

In this work, we apply the technique of categorical logic to these
algebraic structures to design a formal logic, which we call
\emph{ordered predicate logic}, that can be soundly interpreted in
virtual equipments.
%
In ordered predicate logic, the sorts are interpreted as categories,
functions as functors, relations as profunctors and implications of
relations as natural transformations.
%
The type theoretic syntax allows us to write all of our constructions
as if they were ordinary functors and profunctors in set-based
category theory, but the proofs are valid in an arbitrary virtual
equipments and so are valid in enriched and internal category theory
as well.
%
This is analogous to the situation in intuitionistic type theory,
where the syntax is similar to that of informal set theory, but by
restricting to intuitionistic reasoning, the constructions can all be
interpreted in models where types are spaces and terms are continuous
functions ``for free''.

Ordered predicate logic is a kind of \emph{ordered},
\emph{constructive} variant of predicate logic over a unary type
theory.
%
It is \emph{ordered} in that our notion of predicate is that of a
binary relation $R(\alpha;\beta)$ where the $\alpha$ is thought of as
a contravariant dependence and the $\beta$ has a covariant dependence.
%
Proofs in the predicate logic $R_1, R_2,\ldots \vdash S$ are
restricted as in ordered linear logic \cite{lambek-calculus, ordered-linear-logic} in
that each input must be used exactly once and in the order they were
declared.
%
This order discipline ensures that every proof in the logic can be
interpreted as a \emph{natural} transformation.
%
The logic is \emph{constructive} in that we care about the identity of
proofs: they are natural transformations and there may be more than
one natural transformation of the same type.
%
If we ignore the identity of proofs, ordered predicate logic can be
viewed as a logic of preordered sets rather than categories.

The restriction to unary type theory and binary relations is fairly
strict: we cannot discuss important concepts such as functor
categories, opposite categories and arbitrary (co)-ends of
profunctors. However, the restriction is natural in that it
corresponds to a well-understood notion of model (virtual equipments)
and so we leave the extension to future work.
%
We show that a great deal of fundamental results in category theory
are already expressible in this fragment.

Due to the restriction to unary functors and binary profunctors, it is
difficult to state some theorems internally to the theory. For
instance, since we cannot express functor categories, we cannot
express any theorem that quantifies over functors purely in ordered
predicate logic, and instead only as a metatheorem. To more easily
formalize these metatheorems, we add a third layer to the type theory:
an ordinary dependent type theory in the style of Martin-L\"of type
theory. Then all of our ordered predicate logic judgments will also be
indexed by a context from this dependent type theory, and the type
theory includes types for categories, functors, profunctors and
natural transformations that allow us to formalize theorems the logic
is too restrictive to encode.

The dependent type theory layer also has an additional purpose: since
we can quantify over the structures in ordered predicate logic, we can
additionally encode \emph{signatures} for ordered predicate logic as
simply contexts of the dependent type theory. This allows for a clean
formalization of many categorical structures such as monoidal
categories, or strong monads.

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic has been used to develop logics in which
logical relations proofs can be developed such as Abadi-Plotkin logic
for parametricity and LSLR for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr}.
%
Many \emph{binary} logical relations are based on ordered structures,
such as logical relations for approximation in non-deterministic or
partial languages or logical relations for graduality in gradual
typing.
%
We conjecture that some extension of the ordered predicate logic
defined here will be useful for describing these relations as well.

@@

The syntactic forms of Simple Category Theory (SCT) are given in
\cref{fig:syntax}.
%
First, we have the types and terms of the logic over our category
theory. Types include the usual dependent sum, product and identity
types of dependent type theory, as well as several forms that allow us
to abstract over the category theoretic constructions: a type $\Cat$
of categories, a type $\varr{\cat C}{\cat D}$ for functors from $\cat
C$ to $\cat D$, a type $\harr{\cat C}{\cat D}$ for profunctors
contravariant in $\cat C$ and covariant in $\cat D$, and finally a
type $\pendallXdotY {\alpha:{\cat C}} R$ that, when combined with
other constructions on profunctors, allows for abstraction over
arbitrary natural transformations.
%
Terms $L,M,N$ describe the elements of our logic's types, and include
the usual constructions as well as forms for defining categories,
functors, profunctors and natural transformations.
%
Next, we have a syntax for categories, which are written using a
blackboard font $\cat C, \cat D, \cat E$. Categories include terms $M$
of the category type (allowing for abstraction), in addition to
primitives for product and unit categories.
%
Next, we have a syntax for objects of a category, which, because they
are typed with respect to an input variable, provide a syntax for
functors.
%
The objects we can define are object variables, application of a
functor $M a$, and pairs of objects (for the product category) and the
unique object of the unit category.
%
Next, $P,Q,R$ stand for \emph{sets}, but similar to objects, they are
always typed with respect to a contravariant and covariant variable,
and so denote profunctors.
%
These include first the multiplicatives: the hom set
$\punitinXfromYtoZ {\cat C} a b$ representing the morphisms from $a$
to $b$ in $\cat C$, a tensor product $\tensorexistsXwithYandZ
{\beta:\cat D} P Q$, and two kinds of ``ordered function types''
$\homrallXYtoZ \beta P Q$ and $\homlallXYtoZ \alpha R S$. 
%
We also include application of a profunctor term $M$ to two objects,
and cartesian product and unit sets.
%
Next, we have transformations $s,t,u$, which are typed with respect
to (0 or more) input and (one) output set.

Finally, we have three forms of contexts, type contexts $\Gamma$, set
contexts $\Xi$ and transformation contexts $\Phi$. A type context
$\Gamma$ is the familiar notion of context from dependent type
theory. We use uppercase $X,Y,Z$ for term variables to distinguish
them from set variables. Every set expression $P$ must depend on an
object of one category covariantly and an object of one category
contravariantly, however these are allowed to be the same
object. These two scenarios are captured by the notion of set context
$\Xi$. The set context $\alpha:\cat C; \beta: \cat D$ gives the common
case where the contravariant variable $\alpha$ is different from the
covariant variable $\beta$, and the set context $\alpha: \cat C$ alone
gives the case where the profunctor depends on the same variable both
covariantly and contravariantly. Note that we write object variables
as greek letters $\alpha, \beta$. Transformation contexts $\Phi$
describe the inputs to a natural transformation. The input variables
of a natural transformation $x: R$ stand for elements of sets, but as
just described, sets always have covariant and contravariant
dependencies. For this reason, transformation contexts include
category variables $\alpha: \cat C$ interleaved with the set variables
$x : R$, a typical context $\Phi$ would look like $\alpha_1:\cat C_1,
x_1:R_1, \alpha_2: \cat C_2, x_2 : R_2, \alpha_3: \cat C_3$, where
$R_1$ is parameterized by $\alpha_1$ contravariantly and $\alpha_2$
covariantly and $R_2$ is similarly parameterized by
$\alpha_2;\alpha_3$. Note that we consider transformation contexts up
to associativity in that $(\Phi, x:R, \Psi), y : S, \Sigma$ should be
considered equal to $\Phi, x:R, (\Psi, y: S, \Sigma)$.  In this
calculus there is no single ``empty'' transformation context, but
rather the empty context is one that consists of a single category
variable $\alpha:\cat C$ and no set variables. Each form of context
has a corresponding form of substitution, denoted using the lower-case
form of the same greek letters.

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1\\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& M(a,b) \pipe 1 \pipe P \times Q\\
      \textrm{Transformations} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P , \Psi\\
      \textrm{Term Substitution} & \gamma, \delta & ::= & \cdot \pipe \gamma, M/X\\
      \textrm{Object Substitution} & \xi, \zeta & ::= & a/\alpha \pipe a/\alpha;b/\beta\\
      \textrm{Transformation Substitution} & \phi, \psi & ::= & a/\alpha \pipe \phi,s/x, \psi      
    \end{array}
  \end{mathpar}
  \caption{Simple Category Theory Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=TyCtxForm]
        {}
        {\Gamma \isadtctx}

    \inferrule*[right=Cat-form]
    {\Gamma \isadtctx}
    {\Gamma\vdash \cat C \isaCat}

@@

    \inferrule*[right=TransCtxForm]
    {\Gamma \isadtctx}
     {\Gamma \vdash \Phi \isavectx}

    \inferrule*[right=TransFormation]
    {\Gamma \vdash \Phi \isavectx\and
     \Gamma\pipe \underline \Phi \vdash R \isaSet}
    {\Gamma \pipe \Phi \vdash s : R}

    \inferrule
    {\Delta \isadtctx \and \Gamma \isadtctx}
    {\gamma : \Delta \to \Gamma}

    \inferrule*[right=Cat Subst formation]
    {\Gamma \vdash \Psi \isavectx \and \Gamma \vdash \Phi \isavectx}
    {\Gamma \vdash \phi : \Psi \to \Phi}
  \end{mathpar}
  \caption{Formation Rules}
  \label{fig:formation}
\end{figure}

@@

>>>>

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{OHol}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

\newcommand{\paramPresheaf}[1]{\mathcal P^{#1}}
\newcommand{\pmPresheaf}{\paramPresheaf\pm}
\newcommand{\posPresheaf}{\paramPresheaf+}
\newcommand{\negPresheaf}{\paramPresheaf-}
\newcommand{\negPresheafApp}[2]{#1\,#2}
\newcommand{\negPresheafAppPtoX}[2]{#2\in #1}
\newcommand{\posPresheafAppPtoX}[2]{#1 \ni #2}
\newcommand{\posPresheafApp}[2]{#1\,#2}

\newcommand{\graphProf}[3]{\sum_{{#1};{#2}} #3}

@@

\title{A Constructive Ordered Predicate Logic for Category Theory}
\author{Max S. New\inst{1,2} and Daniel R. Licata\inst{2}}
\institute{University of Michigan \and Wesleyan University}

@@

  We develop a sound and complete notion of categorical model for the
  type theory based on virtual equipments, which are known to model
  both internal and enriched category theory. This means that while
  the proofs in our type theory look like standard set-based
  arguments, the syntactic discipline ensure that all proofs and
  constructions carry over to enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that (among many other applications) these
categorical structures were common in logic and computer science, and
that formal systems like logics, type theories and programming
languages can be modeled by sound and complete semantic models given
by notions of structured category\cite{lawvere,goguen,lambek,moggi}, a
phenomenon sometimes referred to as the Curry-Howard-Lambek
nncorrespondence \cite{curry-howard-lambek}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research known as
categorical logic, where new type theories are designed based directly
on category-theoretic semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in pure mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing},
to name a few.

@@

In this work, we apply the technique of categorical logic to these
algebraic structures to design a formal logic, which we call
\emph{virtual equipment type theory}, that can be soundly interpreted
in virtual equipments.
%
In \vett{}, the sorts are interpreted as categories, functions as
functors, relations as profunctors and implications of relations as
profunctor homomorphisms, a generalization of natural transformations.
%
The type theoretic syntax allows us to write all of our constructions
as if they were ordinary functors and profunctors in set-based
category theory, but the proofs are valid in an arbitrary virtual
equipments and so are valid in enriched and internal category theory
as well.
%
This is analogous to the situation in intuitionistic type theory,
where the syntax is similar to that of informal set theory, but by
restricting to intuitionistic reasoning, the constructions can all be
interpreted in models where types are spaces and terms are continuous
functions ``for free''.

\vett{} is a kind of \emph{ordered}, \emph{constructive} variant of
predicate logic over a unary type theory.
%
It is \emph{ordered} in that our notion of predicate is that of a
binary relation $R(\alpha;\beta)$ where the $\alpha$ is thought of as
a contravariant dependence and the $\beta$ has a covariant dependence,
and the rules of the logic distinguish between these two variances.
%
Proofs in the predicate logic $R_1, R_2,\ldots \vdash S$ are
restricted as in ordered linear logic \cite{lambek-calculus, ordered-linear-logic} in
that each input must be used exactly once and in the order they were
declared.
%
This order discipline ensures that every proof in the logic can be
interpreted as a \emph{natural} transformation.
%
The logic is \emph{constructive} in that we care about the identity of
proofs: they are natural transformations and there may be more than
one natural transformation of the same type.
%
If we ignore the identity of proofs, ordered predicate logic can be
viewed as a logic of preordered sets rather than categories.

The restriction to unary type theory and binary relations is fairly
strict: we cannot discuss important concepts such as functor
categories, opposite categories and arbitrary (co)-ends of
profunctors. However, the restriction is natural in that it
corresponds to a well-understood notion of model (virtual equipments)
and so we leave the extension to future work.
%
It is well known that a great deal of fundamental results and
constructions in category theory such as Yoneda and co-Yoneda lemmas,
adjoint functors and Kan extensions are already expressible in virtual
equipments, for instance virtual equipments are used for this purpose
in a text on $\infty$-category theory to avoid explicit manipulation
of model categories \cite{elements-of-oo-category-theory}.
%
We demonstrate some of these standard constructions
in \ref{sec:synthetic} to show that \vett{} gives a syntax in which
these formal constructions look like standard set-theoretic arguments.

Due to the restriction to unary functors and binary profunctors, it is
difficult to state some theorems internally to the theory. For
instance, since we cannot express functor categories, we cannot
express any theorem that quantifies over functors purely in ordered
predicate logic, and instead only as a metatheorem. To more easily
formalize these metatheorems, we add a third layer to the type theory:
an ordinary dependent type theory in the style of Martin-L\"of type
theory. Then all of our ordered predicate logic judgments will also be
indexed by a context from this dependent type theory, and the type
theory includes types for categories, functors, profunctors and
natural transformations that allow us to formalize theorems the logic
is too restrictive to encode. This is analogous to the notion of a
\emph{2-level type theory}: we add a dependent type theory for
reasoning about the terms of our virtual equipment type theory.

The dependent type theory layer also has an additional purpose: since
we can quantify over the structures in \vett{}, we can additionally
encode \emph{signatures} for \vett{} as simply contexts of the
dependent type theory. This allows for a clean formalization of many
categorical structures such as monoidal categories, or strong monads.

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic has been used to develop logics in which
logical relations proofs can be developed such as Abadi-Plotkin logic
for parametricity and LSLR and Iris for step-indexed logical relations
proofs \cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings: operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{ |c|c| } 
      \hline
      Higher-Order Logic & Virtual Equipment Type Theory \\
      \hline
      Set $X$ & Category $\mathbb{C}$\\
      $X \times Y$ & $\mathbb C \times \mathbb D$ \\ 
      1 & $1$ \\
      $\mathcal P X$ & $\mathcal P^+ X, \mathcal P^- X$ \\
      $\{ (x,y) \in X \times Y | R(x,y) \}$ & $\sum_{\alpha:C;\beta:D} R$\\
      Function $f(x:X) : Y$ & Functor/Object $\alpha:\cat C \vdash A : \cat D$\\
      Relation $R(x,y)$ & Profunctor/Set $\alpha:\cat C; \beta:\cat D \vdash R$\\
      $R \wedge Q$ & $R \times Q$\\
      $\top$ & $1$\\
      $\forall x. P \Rightarrow Q$ & $\homrallXYtoZ {\alpha:\cat C} P Q, \homlallXYtoZ {\alpha:\cat C} P Q$ \\
      $\exists x. P \wedge Q$ & $\tensorexistsXwithYandZ {\alpha:\cat C} P Q$\\
      $x =_X y$ & $\punitinXfromYtoZ {\cat C} \alpha \beta$\\
      Proof $\forall \overrightarrow \alpha. R_1(\alpha_1,\alpha_2) \wedge \cdots \Rightarrow Q$ & Nat. Trans./Element $\alpha_1,x_1: R_1(\alpha_1,\alpha_2),\ldots \vdash t : Q$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{Analogy between Higher-Order Logic and \vett{} Judgments and Connectives}
\end{figure}

The syntactic forms of \vett{} are given in
\cref{fig:syntax}.
%
First, we have categories, which are analogous to sorts in a
first-order theory. We have $M$ a base sort, product and unit sorts,
as well as the graph of a profunctor and the negative and positive
presheaf categories.
%
Next, objects $a,b,c$ are the syntax for the \emph{functors} between
categories. We call them objects rather than functors, because in
type-theoretic style, a functor is viewed as a ``generalized object''
parameterized by an input variable $\alpha : \cat C$.
%
Next, sets $P,Q,R$ are the syntax for \emph{sets}. These sets denote
\emph{profunctors}, i.e., a categorification of relations. Similar to
functors, rather than writing profunctors as functions ${\cat
  C}^o\times {\cat D} \to \Set$, we write them as sets with a
contravariant variable $\alpha:\cat C$ and a covariant variable
$\beta:\cat D$. The sets we can define are the Hom-set, the tensor and
internal hom, as well as products of sets, profunctors applied to two
objects and elements of positive and negative presheaves. Finally we
have elements of sets, which correspond to natural transformations of
multiple inputs, where again we view natural transformations valued in
a profunctor as generalized elements of profunctors.

After these forms we have types and terms, which are for the most part
the same as in dependent type theory, except that we have universe
objects for classifying categories, functors, profunctors and natural
transformations.
%
Finally we have several forms of context which are used in the theory:
contexts $\Gamma$ of term variables with their types, contexts $\Xi$
of object variables with their categories, and contexts $\Phi$ of
element variables with their sets, alternating with those sets' object
variables.

%% %
%% First, we have the types and terms of the logic over our category
%% theory. Types include the usual dependent sum, product and identity
%% types of dependent type theory, as well as several forms that allow us
%% to abstract over the category theoretic constructions: a type $\Cat$
%% of categories, a type $\varr{\cat C}{\cat D}$ for functors from $\cat
%% C$ to $\cat D$, a type $\harr{\cat C}{\cat D}$ for profunctors
%% contravariant in $\cat C$ and covariant in $\cat D$, and finally a
%% type $\pendallXdotY {\alpha:{\cat C}} R$ that, when combined with
%% other constructions on profunctors, allows for abstraction over
%% arbitrary natural transformations.
%% %
%% Terms $L,M,N$ describe the elements of our logic's types, and include
%% the usual constructions as well as forms for defining categories,
%% functors, profunctors and natural transformations.
%% %
%% Next, we have a syntax for categories, which are written using a
%% blackboard font $\cat C, \cat D, \cat E$. Categories include terms $M$
%% of the category type (allowing for abstraction), in addition to
%% primitives for product and unit categories.
%% %
%% Next, we have a syntax for objects of a category, which, because they
%% are typed with respect to an input variable, provide a syntax for
%% functors.
%% %
%% The objects we can define are object variables, application of a
%% functor $M a$, and pairs of objects (for the product category) and the
%% unique object of the unit category.
%% %
%% Next, $P,Q,R$ stand for \emph{sets}, but similar to objects, they are
%% always typed with respect to a contravariant and covariant variable,
%% and so denote profunctors.
%% %
%% These include first the multiplicatives: the hom set
%% $\punitinXfromYtoZ {\cat C} a b$ representing the morphisms from $a$
%% to $b$ in $\cat C$, a tensor product $\tensorexistsXwithYandZ
%% {\beta:\cat D} P Q$, and two kinds of ``ordered function types''
%% $\homrallXYtoZ \beta P Q$ and $\homlallXYtoZ \alpha R S$. 
%% %
%% We also include application of a profunctor term $M$ to two objects,
%% and cartesian product and unit sets.
%% %
%% Next, we have transformations $s,t,u$, which are typed with respect
%% to (0 or more) input and (one) output set.

%% Finally, we have three forms of contexts, type contexts $\Gamma$, set
%% contexts $\Xi$ and transformation contexts $\Phi$. A type context
%% $\Gamma$ is the familiar notion of context from dependent type
%% theory. We use uppercase $X,Y,Z$ for term variables to distinguish
%% them from set variables. Every set expression $P$ must depend on an
%% object of one category covariantly and an object of one category
%% contravariantly, however these are allowed to be the same
%% object. These two scenarios are captured by the notion of set context
%% $\Xi$. The set context $\alpha:\cat C; \beta: \cat D$ gives the common
%% case where the contravariant variable $\alpha$ is different from the
%% covariant variable $\beta$, and the set context $\alpha: \cat C$ alone
%% gives the case where the profunctor depends on the same variable both
%% covariantly and contravariantly. Note that we write object variables
%% as greek letters $\alpha, \beta$. Transformation contexts $\Phi$
%% describe the inputs to a natural transformation. The input variables
%% of a natural transformation $x: R$ stand for elements of sets, but as
%% just described, sets always have covariant and contravariant
%% dependencies. For this reason, transformation contexts include
%% category variables $\alpha: \cat C$ interleaved with the set variables
%% $x : R$, a typical context $\Phi$ would look like $\alpha_1:\cat C_1,
%% x_1:R_1, \alpha_2: \cat C_2, x_2 : R_2, \alpha_3: \cat C_3$, where
%% $R_1$ is parameterized by $\alpha_1$ contravariantly and $\alpha_2$
%% covariantly and $R_2$ is similarly parameterized by
%% $\alpha_2;\alpha_3$. Note that we consider transformation contexts up
%% to associativity in that $(\Phi, x:R, \Psi), y : S, \Sigma$ should be
%% considered equal to $\Phi, x:R, (\Psi, y: S, \Sigma)$.  In this
%% calculus there is no single ``empty'' transformation context, but
%% rather the empty context is one that consists of a single category
%% variable $\alpha:\cat C$ and no set variables. Each form of context
%% has a corresponding form of substitution, denoted using the lower-case
%% form of the same greek letters.

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=Cat-form]
    {\Gamma \isadtctx}
    {\Gamma\vdash \cat C \isaCat}

@@

    \inferrule*[right=TransFormation]
    {\Gamma \vdash \Phi \isavectx\and
     \Gamma\pipe \underline \Phi \vdash R \isaSet}
    {\Gamma \pipe \Phi \vdash s : R}
  \end{mathpar}
  \caption{Selected Formation Rules}
  \label{fig:formation}
\end{figure}

@@

\section{Syntactic Details}

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Term Substitution} & \gamma, \delta & ::= & \cdot \pipe \gamma, M/X\\
      \textrm{Object Substitution} & \xi, \zeta & ::= & a/\alpha \pipe a/\alpha;b/\beta\\
      \textrm{Transformation Substitution} & \phi, \psi & ::= & a/\alpha \pipe \phi,s/x, \psi      
    \end{array}

      \inferrule*[right=TyCtxForm]
        {}
        {\Gamma \isadtctx}

    \inferrule*[right=TransCtxForm]
    {\Gamma \isadtctx}
     {\Gamma \vdash \Phi \isavectx}
    \inferrule
    {\Delta \isadtctx \and \Gamma \isadtctx}
    {\gamma : \Delta \to \Gamma}

    \inferrule*[right=Cat Subst formation]
    {\Gamma \vdash \Psi \isavectx \and \Gamma \vdash \Phi \isavectx}
    {\Gamma \vdash \phi : \Psi \to \Phi}

  \end{mathpar}
  \caption{Substitution}

\end{figure}
